Elephant Seal Age JAE Intraspecific Variation

### 2023-12-05 paper/paperv3.qmd
Figure 1 and 2 edits

---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false
    fig-width: 4.5
    fig-height: 4.5
    code-fold: true
    code-tools: true
    warning: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)

@@

```{r}
#| label: age_breeding_mod
#| fig-width: 9

breeding_mod <- glmer(observed_int ~ poly(age, 2) + (1 | animalID) + (1 | year),
                      data = filter(sealdat), 
                      family = "binomial")

@@

summary(breeding_mod)

@@

@@

pup_recruit_model <- glmer(puprecruited ~ poly(age, 2) + (1 | animalID) + (1 | year),
                           data = filter(sealdat, year > 1984 & year < 2019), 
                           family = "binomial")
summary(pup_recruit_model)

pup_surv_recruit <- sealdat %>% 
  mutate(pup_surv = ifelse(pupseeneveragain > 0, 1, 0)) %>% 
  filter(pupsex == "F", 
         pupyearborn < 2016) %>% 
  group_by(age) %>%
  summarize(n_surv = sum(!is.na(pup_surv)),
            mean_surv = sum(pup_surv, na.rm = TRUE) / n_surv,  
            se_surv = (mean_surv * (1 - mean_surv) / n_surv)^0.5,
            n_recruit = sum(!is.na(puprecruited)), 
            mean_recruit = sum(puprecruited, na.rm = TRUE) / n_recruit, 
            se_recruit = (mean_recruit * (1 - mean_recruit) / n_recruit)^0.5)  %>%  
  pivot_longer(cols = c(mean_surv, se_surv, mean_recruit, se_recruit), 
               names_to = c("stat", "metric"), 
               names_pattern = "(.*)_(.*)") %>% 
  pivot_wider(names_from = "stat", 
              values_from = "value") %>% 
  mutate(low = mean - se, 
         high = mean + se, 
         age = age + ifelse(metric == "surv", -0.1, 0.1), 
         metric = factor(metric, 
                         levels = c("recruit", "surv"), 
                         labels = c("Recruitment", "Survival"))) %>% 
  filter(n_surv > 5) %>% 
  ggplot(aes(x = age, y = mean, color = metric)) + 
  geom_pointrange(aes(ymin = low, ymax = high)) +
  geom_ribbon(aes(x),
              pup_recruit_model,
              inherit.aes = FALSE,
              fill = "#829BD1",
              alpha = 0.25) + 
  scale_color_manual(values = c("#829BD1", "#E9B668")) + 
  scale_y_continuous(labels = scales::percent, limits = c(0, .5)) + 
  labs(x = "Maternal Age (Years)", y = "% Pups (Mean ± SE)") +
  theme_few() + 
  theme(legend.position = c(0.1,0.97), 
        legend.justification = c(0,1), 
        legend.title = element_blank(), 
        legend.direction = "horizontal", 
        axis.title = element_text(size = 13)
)

pup_surv_recruit

@@

Results of the Pup Survival model, with individual and year as random effects:

```{r}
#Panel B Survival Model
surv_dat <- sealdat %>% filter(pupyearborn < 2016)
pup_survival_model <- glmer(pup_survived ~ age + (1 | animalID) + (1 | year),
                            data = surv_dat,
                            family = "binomial")
summary(pup_survival_model)

Results of the Pup Recruitment model, with individual and year as random effects:

```{r}
#Panel B Recruitment Model

pup_recruit_model <- glmer(puprecruited ~ age + (1 | animalID) + (1 | year),
                           data = filter(sealdat, year > 1984 & year < 2019), 
                           family = "binomial")
summary(pup_recruit_model)

Modeling the above metrics as quadratic, just in case:

```{r}
surv_dat <- sealdat %>% filter(pupyearborn < 2016) %>% 
  mutate(age10 = age/10)

pup_survival_model <- glmer(pup_survived ~ poly(age, 2) + (1 | animalID) + (1 | year),
                            data = surv_dat,
                            family = "binomial")
summary(pup_survival_model)
#significant result, plot as quadratic on figure

recruit_dat <- sealdat %>% mutate(age10 = age / 10)

pup_recruit_model <- glmer(puprecruited ~ poly(age, 2) + (1 | animalID) + (1 | year),
                           data = filter(recruit_dat, year > 1984 & year < 2019), 
                           family = "binomial", 
                           control = glmerControl(optimizer = "bobyqa"))
summary(pup_recruit_model)
#divide age by 10 (ie rescale variables) - did not work
#increase max iterations within glmer? 
#change the optimizer? -> worked

#survival but not recruitment success with age
#figure with both models, with the recruitment one dashed or something

#ranef variance goes in the table legend

Neither were significant as 2nd order polynomials.

>>>>

---
title: "Elephant Seal Age JAE Intraspecific Variation"
format: 
  html: 
    echo: false
    fig-width: 4.5
    fig-height: 4.5
    code-fold: true
    code-tools: true
    warning: false
    eval: false
editor: visual
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false  
#| eval: true
library(ggthemes)
library(cowplot)
library(lme4)
library(lmerTest) # Loading this library will automatically append p-values to lmer models estimated using something called Satterthwaite's method
library(merTools)
library(tidyverse)
library(ggeffects)
library(patchwork)

@@

```{r}
#| label: age_breeding_mod
#| eval: true

breeding_mod <- glmer(observed_int ~ poly(age, 2) + (1 | animalID) + (1 | year),
                      data = filter(sealdat), 
                      family = "binomial", 
                      control = glmerControl(optimizer = "bobyqa"))

@@

breed_mod_summary <- summary(breeding_mod)

format_signif <- function(x, digits) {
  formatC(signif(x, digits = digits), digits = digits, format = "fg", flag = "#")
}
format_pval <- function(p, digits) {
  ifelse(p < 10^-digits,
         sprintf("<%s", format(10^-digits, digits = 1, scientific = FALSE)),
         sprintf(paste0("%0.", digits, "f"), p))
}

#p values are also rounded....

@@

```{r}
#| tbl-cap: str_glue("The results from breeding probability analyis. The random effect variance of animal id is {VarCorr(breeding_mod, digits = 3)[1]} and year was VarCorr(breeding_mod, digits = 3)[2]".)
#| eval: true

breed_mod_summary$coefficients %>% 
  as_tibble() %>% 
  mutate(Parameter = c("Intercept", "Age (linear)", "Age (quadratic)"),
         across(c(Estimate, `Std. Error`, `z value`), 
                \(x) format_signif(x, digits = 2)),
         `Pr(>|z|)` = format_pval(`Pr(>|z|)`, digits = 3)) %>%
  relocate(Parameter) %>% 
  knitr::kable()
```

@@

## NOTE: REMOVED RANDOM EFFECT OF INDIVIDUAL, OTHERWISE SINGULAR FIT (HYPOTHESIS: 
## INSUFFICIENT SAMPLE SIZES PER INDIVIDUAL)

surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  filter(pupsex == "F",
         pupyearborn < 2016) %>% 
  mutate(pup_survived_dbl = ifelse(pup_survived, 1, 0)) %>% 
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>% 
  filter(n_pups >= 5)
pup_surv_model <- glmer(pup_survived_dbl ~ poly(age, 2) + (1 | year),
                        data = surv_recruit_data, 
                        family = "binomial")
pup_surv_pred <- ggpredict(pup_surv_model, terms = c("age [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x, 
            metric = "surv", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
pup_recruit_model <- glmer(puprecruited ~ poly(age, 2) + (1 | year),
                           data = surv_recruit_data, 
                           family = "binomial")
pup_recruit_pred <- ggpredict(pup_recruit_model, terms = c("age [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x, 
            metric = "recruit", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
surv_recruit_pred <- rbind(pup_surv_pred, pup_recruit_pred) %>% 
  mutate(metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

surv_recruit_summary <- surv_recruit_data %>% 
  group_by(age) %>%
  summarize(n_pups = n(),
            n_surv = sum(pup_survived),
            mean_surv = n_surv / n_pups,  
            se_surv = (mean_surv * (1 - mean_surv) / n_pups)^0.5,
            n_recruit = sum(puprecruited), 
            mean_recruit = n_recruit / n_pups, 
            se_recruit = (mean_recruit * (1 - mean_recruit) / n_pups)^0.5)  %>%  
  pivot_longer(cols = c(mean_surv, se_surv, mean_recruit, se_recruit), 
               names_to = c("stat", "metric"), 
               names_pattern = "(.*)_(.*)") %>% 
  pivot_wider(names_from = "stat", 
              values_from = "value") %>% 
  mutate(low = mean - se, 
         high = mean + se, 
         age = age + ifelse(metric == "surv", -0.1, 0.1), 
         metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

surv_recruit_summary %>% 
  ggplot(aes(x = age, y = mean, color = metric, fill = metric)) + 
  geom_ribbon(aes(ymin = mean_lwr, ymax = mean_upr, alpha = metric),
              surv_recruit_pred,
              color = NA) +
  geom_pointrange(aes(ymin = low, ymax = high)) +
  geom_line(aes(linetype = metric),
            surv_recruit_pred) +
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = c("#829BD1", "#E9B668")) + 
  scale_fill_manual(values = c("#829BD1", "#E9B668")) + 
  scale_linetype_manual(values = c("solid", "dashed")) +
  scale_alpha_manual(values = c(0.25, 0.1)) +
  expand_limits(y = c(0, 0.5)) +
  labs(x = "Maternal Age (Years)", y = "% Pups (Mean ± SE)") +
  theme_few() + 
  theme(legend.position = c(0.1,0.97), 
        legend.justification = c(0,1), 
        legend.title = element_blank(), 
        legend.direction = "horizontal", 
        axis.title = element_text(size = 13))

@@
